Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-11-07 docs/ms_sageAbundance.Rmd
Andys edits to the introduction

Species distribution models (SDMs), on the other hand, are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014].
In a climate change context, SDMs typically rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009].
Thus, interannual dynamics that reflect population responses to yearly climate variations are not represented in SDMs.
Furthermore, since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- in areas where a species currently exists and is expected to persist in the future.
Despite these limitations, SDMs remain one of the few ways to make forecasts of species responses to climate change at large spatial extents.

The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014].
Unfortunately, such methods have been elusive due to the lack of time series data at large spatial extents and the statistical machinary needed to fit high-dimensional spatial models.
Fortunately, two recent technological advances put us in the position to develop large-scale population models.
First, remotely-sensed (RS) time series of species-specific plant cover can provide data on interannual population dynamics that can be correlated with yearly climate fluctuations at large spatial extents.
RS time series have revolutionized models of ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
However, these methods have yet to be used in a dynamic setting.

Building on these technological advances, we develop a statistical approach for modeling plant populations over large spatial extents using a process model motivated by Gompertz density-dependent population growth.
Our approach leverages a unique remotely-sensed time series of *Artemisia* (sagebrush) species [@Homer2012] to fit a spatio-temporal model of plant population dynamics at the landscape scale.
We use advanced dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] to rigorously account for spatial autocorrelation and to produce spatially-explicit estimates of sagebrush percent cover.
We then use the model to forecast the impacts of climate change on sagebrush percent cover.

Sagebrush ecosystems offer an ideal testing ground for our new modeling approach: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
Species distribution models typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss overall in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015].
Sagebrush is thought to depend on winter snow and rain that recharges deeper soil layers [@Pechanec1937; @Schlaepfer2011; @Germino2014].
Models suggest that warming could drive sagebrush losses at low elevations and southern parts of its distribution by making snowmelt earlier and increasing evaporation, leading to soil water deficits in the spring [@Schlaepfer2012; @Schlaepfer2014].
In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Thus, the response of sagebrush to climate will likely vary in space depending on elevation, among other factors.
Direct observations of individual plants and cover on experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season and an increase in sagebrush growth and cover [@Perfors2003; @Harte2015].

>>>>

The ideal tool for forecasting future population states under climate change would be a large-scale dynamic population model [@Schurr2012a; @Merow2014]; however, developing useful models at this scale has been limited by the lack of time series data at large spatial extents and the lack of statistical machinary needed to fit high-dimensional spatial models.
Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
First, new remote-sensing (RS) methods are now producing accurate time series of species-specific plant cover at landscape scales.  These data can be fit with dynamic population models which include yearly fluctuation in climate as covariates.
Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques that overcome the dfficulties of fitting high-dimension spatiotemporal models [see @Conn2015 for a review]. These new statistical methods have yet to be applied to RS derived plant populations data at large scales.

Large-scale, spatially explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes--especially climate change--playing out at landscape and regional scales.  Currently most assesments of how plant and animal populations will respond to climate change rely on Species distribution models (SDMs).  SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014]. However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs. Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [add EhrlenANDMorris]. Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached within our lifetimes. Large-scale dynamic population models could overcome both of these limitations.  They would produce spatially expicit estimates of species abundance within the species range [Ehrlen and Morris] and would provide testable predictions of how populations should respond to short-term climate perturbations.  These short-term predictions would give ecological forecasters the opportunity to repeatedly validate and refine their models (cite a prediction paper).

Sagebrush (*Artemisia tridentata*) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data: sagebrush species are widely distributed [@Kuchler1964], sagebrush species are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].  SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. In contrast, at higher elevations and colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014]. A large-scale spatially explicit population model for sagebrush driven by interannual climate would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.

(Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].) <-- Too much detail? could move to discussion or cite as evidence that "sagebrush is sensitive to climate" above?

Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data, we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future.  Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012]. We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
